---
title: "Covid-19 in Belgium"
description: "How Big Brother is being your Big Brother"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    theme: theme.css
---


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE, 
                      warning=FALSE)
```

```{css}
d-article {
  background-color: #F5FFFA;
}

```


# Introduction

This blog helps understand the COVID-19 numbers and some of the behind the scenes of how policies were made to protect people. 
COVID-19 is a virus that has taken over the world for the last 19 months. It is a virus that causes some kind of respiratory distress. The best way to protect oneself from contracting the virus is by staying away from crowded places, washing your hands or at least sanitizing them.

Belgium is a country in Western Europe. It is a country with a population of 11.59 million.It has a temperate kind of climate. Making the people here less susceptible to the virus as it thrives in colder weather. French, Dutch and German are the three most commonly languages in the country. Currently, the Federal Government exercises federal executive power.

# Data Description

The first data set is from https://ourworldindata.org/covid-deaths. This is data set has 67 variables and provides necessary information about COVID around the globe. After filtering for Belgium I have picked out variables of my choice. Namely : date, new cases, new tests, new vaccinations and new deaths.  For Belgium data was available from February 2020 till August 2021. 
Here the date variable is in date format while the rest are in double format or indicating numbers. 

# Numerical Statistics and what they say:
The numerical Summary statistics very correctly indicate the initial surge of COVID in Belgium between February and May 2020. Then started rising again in August, reaching an all time high in early October which has seemed to be on the decline only form earlier this year.  I am showing the mean which is the average of the numbers, in each of the categories and standard deviation which shows the deviation from the mean.  
The summary statics also show us how many vaccine shots were administered since it became available in December last year.



```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(DT)
library(rmarkdown)
library(formattable)
```

```{r}

covid_data <- read_excel("Data/owid-covid-data.xlsx")

 
Belgium <- covid_data %>% filter(location == "Belgium") %>% 
  select(c("location",
           "date",
           "new_cases",
           "new_deaths",
           "new_tests",
           "new_vaccinations")) %>% 
  mutate(month = month(date, label = TRUE), year = year(date)) %>%
  arrange(year) 

  

Belgium[is.na(Belgium)] <- 0

belgium_table <- Belgium %>% 
  group_by(year, month) %>% 
  summarise(cases = round(mean(new_cases),2),
            deaths= round(mean(new_deaths),2),
            tests= round(mean(new_tests),2),
            vaccinations= round(mean(new_vaccinations),2))
 
 tab1 <- belgium_table %>%
  pivot_longer(cases : vaccinations, names_to = "variables", values_to = "mean")

  tab2 <- Belgium %>% 
    group_by(year, month) %>%
  summarise(cases = round(sd(new_cases),2),
            deaths= round(sd(new_deaths),2),
            tests= round(sd(new_tests),2),
            vaccinations= round(sd(new_vaccinations),2)) %>%
    pivot_longer(cases : vaccinations, names_to = "variables", values_to = "sd")



bel <- left_join(tab1, tab2)




bel %>%  DT::datatable(rownames = FALSE, 
                                 colnames = c("Year", "Month", "variables", "mean", "sd"),
                                 options = list(pageLength = 8))


```


## How did big brother help?

- Big brother is a famous term coined in the book 1984 by George Orwell. He used to indicate how digital technology is being used to track the world's every move. But in a pandemic this could actually be used for the right reasons. 

- For instance, every persons location is being tracked using their cellular device. So the same anonymised insights used in Google maps was helpful in making critical decisions to combat the virus.  

- This data, shows how visits to places like, grocery stores and parks have changed when compared to a baseline. A baseline is the median value for the corresponding day of the week. 

- It has this information for 135 countries and even a state wise break up for some. This data was not used for medical diagnostic or prognostic purposes. It can help set working hours more effectively to avoid over crowding, set the requirement, timing and frequency of the transport at transit stations. This also help understanding the destination of the travelers and provide guidance to protect public health and needs of essential commodities.  



```{r}

google <- read_excel("Data/Global_Mobility_Report.xlsx")

google_clean <- google %>% 
  filter(country_region == "Belgium") %>% 
mutate(month = month(date, label = TRUE), year = year(date)) %>%
  arrange(year) %>% 
select(c("country_region",
           "month",
           "year",
           "retail_and_recreation_percent_change_from_baseline",
           "grocery_and_pharmacy_percent_change_from_baseline",
           "parks_percent_change_from_baseline",
           "transit_stations_percent_change_from_baseline",
           "workplaces_percent_change_from_baseline"
          )) %>% 
  
  rename(retail_rec_ = retail_and_recreation_percent_change_from_baseline,
            grocery_pharma = grocery_and_pharmacy_percent_change_from_baseline,
            parks = parks_percent_change_from_baseline,
            transit_stations = transit_stations_percent_change_from_baseline,
            work = workplaces_percent_change_from_baseline) 

 google_clean[is.na(google_clean)] <- 0

```


```{r}
google_table <- google_clean %>% 
  group_by(year, month) %>% 
  summarise(retail_rec_ = round(mean(retail_rec_),2),
            grocery_pharma = round(mean(grocery_pharma),2), 
            parks = round(mean(parks),2),
            transit_stations = round(mean(transit_stations),2),
            work = round(mean(work),2)) 
            
 

```

```{r}

google_table %>% 
  select(year, month, retail_rec_, grocery_pharma, parks, transit_stations, work) %>% 
  as.data.frame() %>% 
  rownames_to_column("S.no") %>% 
  formattable(caption = "Percentage of mobility from baseline", list(area(col = 3:8) ~ color_tile( "#32CD32", "#FF0000")))

```



1. From this it is obvious that the population staying at home and away form the work place has increased. Similarly transit stations (bus, metro and train stations) have also seen a large drop in people visiting them. 

2. Retail stores and recreational areas  initially saw a huge drop and then around June last year has seen more movement. Then again from the resurgence in October last year it saw a drop in population with it increasing again in last couple of months. 

3. Pharmacies however saw the least impact of a lock down. For obvious reason they had seen a more influx of movement compared to any other category of places considered barring Parks. 

4. Parks however, in the minds of people, is indicated as an open space with enough area for social distancing and hence saw the maximum movement over the last 19 months. 

From this data it can be understood, while a good portion of the population stayed at home and away from places of work, there was still a certain amount of restricted movement in the country. 

# Conclusion:

Belgium has seen a record number of deaths despite it's population. Using the above tables we can find out that even during the months where the cases were very high like April and October 2020 the flux of mobility among the populaiton was still quite high with more then 20% of the population not following the restrictions at any place during the pandemic. 

# Citations:

## Packages Used:

 Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with lubridate. Journal of Statistical Software, 40(3), 1-25. URL
  https://www.jstatsoft.org/v40/i03/.

 Hadley Wickham and Jennifer Bryan (2019). readxl: Read Excel Files. R package version 1.3.1. https://CRAN.R-project.org/package=readxl

 JJ Allaire and Yihui Xie and Jonathan McPherson and Javier Luraschi and Kevin Ushey and Aron Atkins and Hadley Wickham and Joe Cheng
  and Winston Chang and Richard Iannone (2021). rmarkdown: Dynamic Documents for R. R package version 2.9. URL
  https://rmarkdown.rstudio.com.

 Kun Ren and Kenton Russell (2021). formattable: Create 'Formattable' Data Structures. R package version 0.2.1.
  https://CRAN.R-project.org/package=formattable

 Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686 

 Yihui Xie, Joe Cheng and Xianying Tan (2021). DT: A Wrapper of the JavaScript Library 'DataTables'. R package version 0.18.
  https://CRAN.R-project.org/package=DT
  
  
## Data:


 COVID-19 Community Mobility Report. 2021. COVID-19 Community Mobility Report. <https://www.google.com/covid19/mobility/> 

 Ritchie, H., Mathieu, E., Rodés-Guirao, L., Appel, C., Giattino, C., Ortiz-Ospina, E., Hasell, J., Macdonald, B., Beltekian, D. and Roser, M., 2021.     Coronavirus Pandemic (COVID-19).  <https://ourworldindata.org/covid-deaths> 

  
## References:

 Google.com. 2021.  <https://www.google.com/covid19/mobility/data_documentation.html?hl=en> 
 
 
 https://knowyourmeme.com/photos/1514752-google-doodles. 2021. <https://knowyourmeme.com/photos/1514752-google-doodles> 


  Mampatta, S., 2021. Explained: Why the Google Mobility Reports matter to policymakers.  <https://www.business-standard.com/article/companies/explained-why-the-google-mobility-reports-matter-to-policymakers-120070600808_1.html>
  
  Who.int. 2021. Coronavirus. <https://www.who.int/health-topics/coronavirus#tab=tab_1>


 
